---
title: "dartR Workshop - CBA"
author: "Bernd Gruber & Arthur Georges"
date: '`r Sys.Date()`'
output: 

#  pdf_document:
#  toc: yes
#    number_sections: true
#    includes: 
#      in_header: boxcba.tex
#  tufte::tufte_html:
  tufte::tufte_handout:
    
#citation_package: natbib
   latex_engine: xelatex
   toc: yes
   

   highlight: tango
   includes:
     in_header: boxcba.tex
     
always_allow_html: yes
        # tufte::tufte_book:
#   citation_package: natbib
#   latex_engine: xelatex
#   highlight: tango
#   includes:
#           in_header: box2.tex
#bibliography: skeleton.bib
#link-citations: yes
---

```{r setup, include=FALSE, }
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = TRUE)
library(tufte)
library(knitr)
library(TeachBayes)
#knitr::opts_chunk$set(prompt = TRUE)
```

\newpage


#Foreword

What it this tutorial about?

In this workshop we provide an overview on the use of a recently developed R-package (dartR) that aims to integrate as many as possible ways to analyse SNP data sets. 
^[A general overview on the package (though quite a few new function have been added can be found here: [Gruber et al. 2018: https://onlinelibrary.wiley.com/ doi/full/ 10.1111/1755-0998.12745](https://onlinelibrary.wiley.com/doi/full/10.1111/1755-0998.12745)].

Below here you will find exercises that should get you familiar with the package. Feel free to use the provided sample data or use you own data.




# Landscape genetics 

This tutorial is slightly different from the previous ones as it will guide you through an example and explains you how to create a simple resistance layer. Then you are encouraged to explore the approach using your own data or the example data set provided.


##Guided example: A landsape genetic analysis
    
The idea of a landscape genetic analysis is that genetic similarity is between individuals/populations is dependent on the distance between individuals and [potentially] on the "resistance" of the landscape between individuals/populations. 

For this example we first load a data set called possums, which is already in genlight format and comes with the packages.

```{r, message=FALSE, warning=FALSE}
library(dartR)
```


```{r}
possums.gl
```

```{block, type="task"}

1. Study the possum genlight object (how many individuals per population)
2. Overall how many loci are  in the data set?

```


## Isolation by distance

The "null model" in landscape genetics is that there is a simple relationship between genetic and euclidean distance. A standard procedure is to study the relationship between log(euclidean distance) and Fst/1-Fst (see ?gl.ibd for details). For a quick check we can use the ```gl.ibd```. To be able to use the function the genlight object needs to have the coordinates for each individual in the ```@other$latlong``` slot. Further we need to provide information if the coordinates are already projected or given in lat/lon.


```{r}

iso <- gl.ibd(possums.gl, projected = TRUE)

```


The function returns a list of the following components: Dgen (the genetic distance matrix), Dgeo (the Euclidean distance matrix), mantel (the statistics of the mantel test)

A mantel test is basically a simple regression, but the significance takes the non-independence of pairwise distances via a bootstrap approach into account.


## Landscape genetics using a landsacpe restistance approach

Often ecologists want to know if a particular landscape feature is affecting population structure on top of Euclidean distance. The idea is that a particular feature is causing some cost for individuals when moving through it, hence modifying the actual euclidean distance between individuals/populations. Commonly used approaches to calculate so-called cost-distances are the "least-cost" and "circuitscape" approach. 
Both approaches require a landscape that represents landscape features in terms of the "resistance" values

## Calculation of cost distances

In this example we use populations as the entity of interest. Hence we need to calculate three distance matrices, namely a Euclidean distance matrix, a cost distance matrix and finally a genetic distance matrix. The two distance matrices can then be used (very similar to the partial mantel test above) to compete against each other how well they explain genetic distances. As mentioned we base our analysis on indviduals, therefore we first need to calculate the coordinates of our population centers. But first we load our (resistance) landscape.


```{r, echo=F}

landscape <- readRDS("./data/landscape.rdata")

```

```{r, eval=F}
landscape <- readRDS(("/data/scratch/GSA/landscape.rdata"))
```

We calculate the population centers via:

```{r, fig.height=5}

xs <- tapply(possums@other$latlong[,"lon"], pop(possums), mean)
ys <- tapply(possums@other$latlong[,"lat"], pop(possums), mean)

plot(landscape)
points(xs, ys, pch=16)
text(xs, ys, popNames(possums), col="orange")

coords <- cbind(xs, ys)

```

## Euclidean distance
```{r}
eucl <- as.matrix(dist(coords))
```


## Costdistances
```{r}
library(PopGenReport)
cost <- costdistances(landscape = landscape, locs = coords, method = "leastcost", NN=8)
```


## genetic distance
For simplicity we will use pairwise Fsts between population here


```{r}
library(StAMPP)
gd <-as.matrix(as.dist(stamppFst(possums, nboots=1)))
```


And finally run a partial mantel test

```{r}
wassermann(gen.mat = gd, eucl.mat = eucl, cost.mats = list(cost=cost),plot = F)
```


Library PopGenReport has a convinience function that does all in once, but is less flexible Please note we need to transform the possums genlight to a genind object.
It has the benefit that it shows the actual least cost path in the landscape (but runs longer).


```{r, fig.height=5}

pgi <-gl2gi(possums)
glc <- genleastcost(pgi, fric.raster = landscape, gen.distance = "D", NN = 8, pathtype = "leastcost")

```


